How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-04-15 doc/paper.tex
additional word smithing to methods and supplement

\begin{abstract}
  A tenet of conservation paleobiology is that knowledge of past extinction patterns can help us to better predict future extinctions. Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past. To answer this question we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton.) For each group we examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state. Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate state) the improvement in predictive power over models with constant effects and no historical covariates is very minor. This model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. It performs nearly as well at out-of-sample prediction (70-80\%), implying that determinants of extinction risk have varied only modestly through time. An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.

  %risk assessments for extant species can be improved by examining past extinction patterns and by using paleontological records to establish geographic range and abundance trajectories on geological timescales. 
  %We ask if, at a given point in time, how accurate are extinction risk predictions based on instantaneous geographic ranges and knowledge of past extinction/survival patterns?
  %Such suggestions prompt two key questions: (1) At a given timescale, to what extent are extinction risk trajectories deterministic (past trends are likely to continue into the future) versus Markovian (the future depends only on the present state)? (2) 
  %Our results imply that at million-year timescales extinction risk trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. 
  %However, models based on present geographic ranges and past range-risk relationships are likely to be moderately successful at predicting future extinction/survival outcomes. 
  %In-sample model performance measures are approximately equal to out-of-sample performance measures (assessed with 5-fold cross-validation), suggesting that the determinants of relative extinction risk are fairly consistent throughout the Cenozoic. 
  %Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future. 
  %We have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past because our in-sample model performance measures are approximately equal to out-of-sample performance. 
  %Including the historical covariates and allowing their effects to vary over time yields marginally better predictions than not including them. 
  %However, the improvement in predictive power by including these historical covariates is modest at best, absent at worst, and ultimately reflects the extremely stochastic nature of species extinction. 
  %Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups. 
  %This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton. 
  %The greatest source of variance in survival probability is the timing of observation. 
  %Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age. 
  %We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. 
  %The effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range. 
  %These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. 
  %The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.  
  %These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small. 
  %The results of this study reinforce the importance of the promise of paleontology and using the past to predict the future.
\end{abstract}

{\bf Keywords:} conservation, paleobiology, extinction, forecasting

@@

Past paleobiological studies of extinction have frequently focused on identifying and measuring the effect of various potential predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}.

@@

We analyzed microfossil occurrence information which was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}. Occurrence information was downloaded for calcareous nannoplankton, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya). 
This dataset of occurrences was then filtered to just those species which have their first occurrence at most 63 Mya. This choice means that our analysis avoids those taxa which survived the K/Pg boundary, those taxa which arose just after the K/Pg boundary, and means that our occurrence histories line up with the temperature time-series which was used as a predictor of extinction as discussed below.

@@

We also calculated the average global temperature of each 1 My bin as estimated from Magnesium/Calcium isotope ratios \citet{Cramer2011}. These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself. This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, \citet{Cramer2011}, provides temperature estimates for every 0.1 My from 0 to 63 Mya. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.

See Section \ref{sec:data_desc} for a further explanation on how observations were temporally binned and how our covariates were standardized and transformed prior to analysis.

@@

We used a discrete-time survival modelling framework in order to estimate how well we can predict extinction risk at one million year time scales. At its core, our model is a multilevel logistic regression with taxon age in millions of years as a varying intercept \citep{Tutz2016}. We considered four model variations of the predictors included in our model: covariate effects are constant over time and none of our historical covariates are included (Model C), covariate effects are allowed to vary over time but none of our historical covariates are included (Model V), covariate effects are constant over time and our historical covariates are included (Model CP), covariate effects are allowed to vary over time and historical covariates are included (Model VP). See Table \ref{tab:model_def} for further explanation of how these models differ from each other. For a complete description of the statistical model used in this analysis, please see Section \ref{sec:model_desc}. For a full description of how our models were implemented, please see Section \ref{sec:model_est}.

Importantly, we include species geologic age at time of observation as a non-nested varying intercept term. This factor may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}, but its inclusion in our model is critical to its nature as a survival model \cite{Tutz2016}. Again, see Section \ref{sec:model_desc} for how this term was specified as part of our model.

@@

In-sample forecasting means we are estimating how well our model predicts extinction probability for observations that that model was fit to. This is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data.

We are particularly interested in understanding how well our model forecasts extinction probability of data from the future that the model was not fit to (out-of-sample data). In-sample forecasting measures, however, are not necessarily good estimates of the model's ability to predict data from the future \citep{ESL}. To quantify our ability to forecast species' extinction risk, we estimated average out-of-sample forecasting performance using 5-fold time-series cross-validation. For time-series data, the folds (data partitions) are approximately equal segments of time. Each fold represents a sequence of time points. With 63 time points, each of the five folds represents approximately 13 time points. Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.

k-fold cross-validation for time series precedes as follows \citep{Arlot2010,Bergmeir2016}. First, the model is fit to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e. the future). Then the model is fit to the combined first and second folds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold. This process continues with the model fit to the first three folds combined is then used to forecast extinction probabilities for the fourth fold. Finally, the model is fit to the first four folds combined and then is used to forecast the fifth fold. The results from these forecasts are then combined to yield our estimate of expected out-of-sample performance.

The relative adequacy of our four model variants were compared using the area under the receiver operating characteristic curve or AUC \citep{Fawcett2006,Mason2002}. This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers true positives. AUC ranges between 0.5 and 1, with 0.5 indicating no difference in classification from random and 1 indicating perfect classification. AUC can be interpreted as the probability that our model correctly ranks the relative extinction risks of a randomly selected extinct-extant species pair \citep{Fawcett2006,Mason2002}. AUC values of approximately 0.8 or greater can be considered ``good'' \citep{ACCDA}, so we interpret values between 0.7 and 0.8 could then be considered ``fair,'' and values between 0.6 and 0.7 as ``poor.''

@@

We find that all of our models are expected to correctly forecasting the rank order in extinction risk of two randomly selected future observations approximately 70\% to 80\% of the time. One of the most striking aspects of this results is that the in-sample and out-of-sample forecasts made with our models demonstrate that while models where the historical covariates are predictors of species extinction have better in-sample performance (Fig. \ref{fig:auc_hist}, \ref{fig:auc_taxon_time}), all of them forecast the extinction risk of out-of-sample data with a similar degree of success (Fig. \ref{fig:fold_auc}, \ref{fig:fold_auc_taxon_time}).

We also find that our four models are practically identical in their ability to make forecasts both in-sample and out-of-sample. Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). While the ``best'' model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance versus the other models is negligible (Fig. \ref{fig:fold_auc}). This means that even though model VP has a statistically greater AUC than the other three models, this result is not practically or scientifically significant.

@@

The quality and consistency of our models out-of-sample forecasting performance is suprising given that its estimates are based on rather limited information about the taxa themselves. Our most complex models are only taking into account some aspects of species geographic range and their taxonomic grouping. The principal reason we were not able to include more biological information in our models is that we lack almost any life history or ecological information on most marine micro- and nanno-plankton. Forams are the exception to this problem -- there is life history, ecological, and physiological information for a selection of foram species \citep{Ezard2011}. However, not this information does not exist all foram species. If we want to include this type of information in a predictive model of extinction risk, we would be able to analyze only a single taxonomic subset of the fossil occurrences present the Neptune Database and then only a limited selection of those species. This means ignoring the majority of occurrence information present in the Neptune database.

@@

Temperature was also transformed and standardized the in the same manner as geographic range. Difference in temperature between observations was measured from the standardized temperature values and was not standardized separately.

@@

In survival analysis, the hazard function describes the instantaneous rate of extinction of a species given its age and relevant covariates. The hazard function is defined as the conditional probability of a species going extinct by the end of the \(t\)-th interval given that it survived up until \(t\) and the relevant covariate information \(X\) for all \(k\) 1 My intervals \citep{Tutz2016}. For the discrete time intervals \(T = 1, \cdots, k\), extinction is defined as \(T = t\). The discrete time hazard function is defined as
\begin{equation}
  \lambda(t | X) = P(T = t | T \geq t, X), \quad t = 1, \cdots, k.
  \label{eq:hazard}
\end{equation}

The hazard function (Eq. \ref{eq:hazard}) is easily reparameterized as a logistic regression by defining that \(\lambda(t | X) = h(\Theta)\) where \(h(.)\) is a logit inverse-link function and \(\Theta\) is the probability of a taxon going extinction during interval \(t\) \citep{Tutz2016}. \(h(\Theta)\) is then modeled as with any regression as it is defined for all real-values. In this case, we opted for a hierarchical/mixed-effects model with multiple non-nested varying intercepts and slopes \citep{ARM}.

@@

We implemented our model (Eq. \ref{eq:model} using the \begin{texttt}rstanarm\end{texttt} package for the R programming language \citep{StanManual}. This package provides an interface to the Stan probabilistic programming language for writing hierarchical/mixed-effects models in native R. Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling. In order to prevent divergent transitions adapt delta was increased to 0.9999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm 2.18.2 \citep{rstanarm}.

An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as:
\begin{verbatim}
form <- event ~ range + range_diff + temp + temp_lag + 
                (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
                (1 | age/phylum), 
stan_glmer(formula = form,
           data = data, 
           family = 'binomial',
           prior = normal(c(-1, 0, 0, 0), rep(1, 4), autoscale = FALSE), 
           prior_intercept = normal(-2, 5, autoscale = FALSE), 
           prior_aux = cauchy(0, 1, autoscale = FALSE), 
           chains = 4,
           thin = 4,
           adapt_delta = 0.999999)
\end{verbatim}

Similarly, our full model can also be implemented using the \begin{texttt}brms\end{texttt} Stan interface \citep{brms2017,brms2018} as:
\begin{verbatim}
priors <- c(set_prior('normal(-2, 5)', class= 'Intercept'),
            set_prior('normal(0, 1)', class = 'b'),
            set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
            set_prior('cauchy(0, 1)', class = 'sd'),
            set_prior('lkj(1)', class = 'cor'))
form <- bf(event ~ range + range_diff + temp + temp_lag +
           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
           (1 | age/phylum))
brmfit <- brm(formula = form,
              data = data, 
              family = bernoulli(), 
              prior = priors,
              chains = 4, 
              thin = 4,
              control = list(adapt_delta = 0.999999)
\end{verbatim}

>>>>

\begin{abstract}
  A tenet of conservation palaeobiology is that knowledge of past extinction patterns can help us to better predict future extinctions. Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past. To answer this question we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton.) For each group we examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state. Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate state) the improvement in predictive power over models with constant effects and no historical covariates is very minor. This model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. It performs nearly as well at out-of-sample prediction (70-80\%), implying that determinants of extinction risk have varied only modestly through time. An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.

  %risk assessments for extant species can be improved by examining past extinction patterns and by using palaeontological records to establish geographic range and abundance trajectories on geological timescales. 
  %We ask if, at a given point in time, how accurate are extinction risk predictions based on instantaneous geographic ranges and knowledge of past extinction/survival patterns?
  %Such suggestions prompt two key questions: (1) At a given timescale, to what extent are extinction risk trajectories deterministic (past trends are likely to continue into the future) versus Markovian (the future depends only on the present state)? (2) 
  %Our results imply that at million-year timescales extinction risk trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. 
  %However, models based on present geographic ranges and past range-risk relationships are likely to be moderately successful at predicting future extinction/survival outcomes. 
  %In-sample model performance measures are approximately equal to out-of-sample performance measures (assessed with 5-fold cross-validation), suggesting that the determinants of relative extinction risk are fairly consistent throughout the Cenozoic. 
  %Our best supported model includes the historical covariates, change in geographic range and lag of global temperature, which indicates that the past improves our estimates of the present and future. 
  %We have confidence that our conclusions about our ability to predict species extinction risk in the future with similar accuracy to when we predict species extinction risk in the past because our in-sample model performance measures are approximately equal to out-of-sample performance. 
  %Including the historical covariates and allowing their effects to vary over time yields marginally better predictions than not including them. 
  %However, the improvement in predictive power by including these historical covariates is modest at best, absent at worst, and ultimately reflects the extremely stochastic nature of species extinction. 
  %Finally, we find support for species' extinction risk increases with age, though the strength of this effect varies among taxonomic groups. 
  %This effect is most pronounced in forams and radiolarians, and less pronounced in diatoms and calcareous nanoplankton. 
  %The greatest source of variance in survival probability is the timing of observation. 
  %Importantly, this result means that the time of an observation is a greater source of variation in survival probability when compared to species age. 
  %We find that including information on a species' change in geographic range size on average improves our predictions of species survival at million-year timescales. 
  %The effect of change in geographic range is much smaller than the effect of current geographic range, and highly variable through time as the effect changes sign and there are times where there is little evidence for any effect of past geographic range. 
  %These results imply that at million-year timescales geographic range trajectories are nearly Markovian, perhaps because the processes driving geographic range changes vary on substantially shorter timescales. 
  %The effect of change in geographic range on survival most likely stands for many interacting and unobserved processes which in-turn produce that species' geographic range and its affect on survival.  
  %These results reflect the difficulty of estimating species extinction, and that while including historical covariates does improve model performance, that gain is very small. 
  %The results of this study reinforce the importance of the promise of palaeontology and using the past to predict the future.
\end{abstract}

{\bf Keywords:} conservation, palaeobiology, extinction, forecasting

@@

Past palaeobiological studies of extinction have frequently focused on identifying and measuring the effect of various potential predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}.

@@

We analyzed microfossil occurrence information from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}. This dataset of occurrences covered calcareous nannoplankton, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya). 
These occurrences were then filtered to include only those species who's first occurrence was no more than 63 Mya. This filtering criterion means that our analysis avoids those taxa which survived the K/Pg boundary, those taxa which arose just after the K/Pg boundary, and ensures that our occurrence histories are congruent with the temperature time-series used as a predictor of extinction (see below).

@@

Average global temperature of each 1 My bin was calculated from estimates based on Magnesium/Calcium isotope ratios \citet{Cramer2011}. Elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself. This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, \citet{Cramer2011}, estimated temperature for every 0.1 My from 0 to 63 Mya. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.

See Section \ref{sec:data_desc} for a further explanation on how observations were temporally binned, and how our covariates were standardized and transformed prior to analysis.

@@

We used a discrete-time survival modelling framework in order to estimate how well we can predict extinction risk at one million year time scales. At its core, our model is a multilevel logistic regression with taxon age in millions of years as a varying intercept \citep{Tutz2016}. We considered four variations of the predictors included in our model: covariate effects are constant over time and none of our historical covariates are included (Model C), covariate effects are allowed to vary over time but none of our historical covariates are included (Model V), covariate effects are constant over time and our historical covariates are included (Model CP), covariate effects are allowed to vary over time and historical covariates are included (Model VP). We always included species geologic age at time of observation as a non-nested varying intercept term. This factor may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}, but its inclusion in our model is critical to its nature as a survival model \cite{Tutz2016}.

See Table \ref{tab:model_def} for further explanation of how the four models we considered differ from each other. A complete description of the statistical model used in this analysis is available in Section \ref{sec:model_desc}. Additionally, the full description of how these models were implemented and coded, including choice of priors, is available Section \ref{sec:model_est}.

@@

In-sample forecasting means we are estimating how well our model predicts extinction probability for observations that that model was fit to. This is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data. In-sample forecasting measures, however, are not necessarily good estimates of the model's ability to predict data from the future \citep{ESL}.

We are particularly interested in understanding how well our model forecasts extinction probability of data from the future that the model was not fit to (out-of-sample data). To quantify our ability to forecast species' extinction risk, we estimated average out-of-sample forecasting performance using 5-fold time-series cross-validation. For time-series data, the folds (data partitions) are approximately equal segments of time. Each fold represents a sequence of time points. With 63 time points, each of the five folds represents approximately 13 time points. Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.

k-fold cross-validation for time series follows a specific sequence of procedures \citep{Arlot2010,Bergmeir2016}. First, the model is fit to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e. the future). Then the model is fit to the combined first and second folds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold. Continuing, the model is then fit to the first three folds combined and is then used to forecast extinction probabilities for the fourth fold. Finally, the model is fit to the first four folds combined and then is used to forecast the fifth fold. The results from these forecasts are then combined to yield our estimate of expected out-of-sample performance.

The relative adequacy of the four model variants was compared using the area under the receiver operating characteristic curve or AUC \citep{Fawcett2006,Mason2002}. This measure is commonly used to measure the performance of classification models as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers true positives. AUC ranges between 0.5 and 1, with 0.5 indicating no difference in classification from random and 1 indicating perfect classification. AUC can be interpreted as the probability that our model correctly ranks the relative extinction risks of a randomly selected extinct-extant species pair \citep{Fawcett2006,Mason2002}. AUC values of approximately 0.8 or greater can be considered ``good'' \citep{ACCDA}, so we interpret values between 0.7 and 0.8 could then be considered ``fair,'' and values between 0.6 and 0.7 as ``poor.''

@@

We find that all of our models are expected to correctly forecasting the rank order in extinction risk of two randomly selected future observations between 70\% to 80\% of the time. One of the most striking aspects of these results is the similarity in forecasting performance for in-sample and out-of-sample observations. A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original). Additionally, a potential decrease in precision when forecasting the future of extinction risk is almost to be expected as the future is not necessarily the same as the past. However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to how extinction intensity has changed over the Cenozoic. An important caveat to our results is that human impacts may substantially alter extinction risk dynamics away from the average Cenozoic ``state'' so that the future could become less predictable than it has been in the past.

%This similarity demonstrates that while models with historical covariates are predictors of species extinction have better in-sample performance (Fig. \ref{fig:auc_hist}, \ref{fig:auc_taxon_time}), all of them forecast the extinction risk of out-of-sample data with a similar degree of success (Fig. \ref{fig:fold_auc}, \ref{fig:fold_auc_taxon_time}).

We also find that our four models are practically identical in their ability to make forecasts both in-sample and out-of-sample. Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). While the ``best'' model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance versus the other models is negligible (Fig. \ref{fig:fold_auc}). This means that even though model VP has a statistically greater AUC for in-sample forecasts than the other three models, this result is not practically or scientifically significant. The results of the out-of-sample forecasts illustrate this point in full, as all four models have functionally identical out-of-sample performance.

@@

The quality and consistency of our models out-of-sample forecasting performance is surprising given that its estimates are based on rather limited biological and environmental information about the taxa. Our most complex models are only taking into account some aspects of species geographic range and their taxonomic grouping. The principal reason we were not able to include more biological information in our models is that we lack almost any life history or ecological information on most marine micro- and nannoplankton. Forams are an exception to this problem as aspects of life history, ecology, and physiology for many foram species \citep{Ezard2011}. However, comparable information does not exist all foram species, nor does this type of data exist for the other three taxonomic groups studied here. Including this type of information in a predictive model of extinction risk we require analyzing a single, limited taxonomic subset of the fossil occurrences present the Neptune Database. This would mean ignoring the majority of occurrence information present in the Neptune database.

%It performs nearly as well at out-of-sample prediction (70-80\%), implying that determinants of extinction risk have varied only modestly through time. An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.

@@

Temperature was also transformed and standardized the in the same manner as geographic range. The change in temperature between an observation and its previous observation was measured from the standardized temperature values and was not standardized separately.

@@

In survival analysis, the hazard function describes the instantaneous rate of extinction of a species given its age and covariate information. The hazard function is defined as the conditional probability of a species going extinct by the end of the \(t\)-th interval given that it survived up until \(t\) and the relevant covariate information \(X\) for all \(k\) 1 My intervals \citep{Tutz2016}. For the discrete time intervals \(T = 1, \cdots, k\), extinction is defined as \(T = t\). The discrete time hazard function is defined as
\begin{equation}
  \lambda(t | X) = P(T = t | T \geq t, X), \quad t = 1, \cdots, k.
  \label{eq:hazard}
\end{equation}

The hazard function (Eq. \ref{eq:hazard}) is easily reparameterized as a logistic regression by defining that \(\lambda(t | X) = h(\Theta)\) where \(h(.)\) is a logit inverse-link function and \(\Theta\) is the probability of a taxon going extinction during interval \(t\) \citep{Tutz2016}. \(h(\Theta)\) is then modeled as with any regression. In this case, we opted for a hierarchical/mixed-effects model with multiple non-nested varying intercepts and slopes \citep{ARM}.

@@

We implemented our model (Eq. \ref{eq:model} using the \begin{texttt}rstanarm\end{texttt} package for the R programming language \citep{StanManual}. This package provides an interface to the Stan probabilistic programming language for writing hierarchical/mixed-effects models in native R. Posterior estimates were obtained through Hamiltonian Monte Carlo, using 2000 steps divided equally between warm-up and sampling. In order to prevent divergent transitions adapt delta was increased to 0.999999; all other HMC/NUTS sampling parameters were kept at the defaults for rstanarm 2.18.2 \citep{rstanarm}.

To implement our VP model in \begin{texttt}rstanarm\end{texttt}, where ``data'' is a data.frame object of all necessary data (response, covariates), is coded as:
\begin{verbatim}
form <- event ~ range + range_diff + temp + temp_lag + 
                (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
                (1 | age/phylum), 
stan_glmer(formula = form,
           data = data, 
           family = 'binomial',
           prior = normal(c(-1, 0, 0, 0), rep(1, 4), autoscale = FALSE), 
           prior_intercept = normal(-2, 5, autoscale = FALSE), 
           prior_aux = cauchy(0, 1, autoscale = FALSE), 
           chains = 4,
           thin = 4,
           adapt_delta = 0.999999)
\end{verbatim}

Similarly, our VP model can be implemented using the \begin{texttt}brms\end{texttt} Stan interface \citep{brms2017,brms2018} as:
\begin{verbatim}
priors <- c(set_prior('normal(-2, 5)', class= 'Intercept'),
            set_prior('normal(0, 1)', class = 'b'),
            set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
            set_prior('cauchy(0, 1)', class = 'sd'),
            set_prior('lkj(1)', class = 'cor'))
form <- bf(event ~ range + range_diff + temp + temp_lag +
           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
           (1 | age/phylum))
brmfit <- brm(formula = form,
              data = data, 
              family = bernoulli(), 
              prior = priors,
              chains = 4, 
              thin = 4,
              control = list(adapt_delta = 0.999999)
\end{verbatim}
